---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Init
```{r}
library(ggplot2)
library(scales)
time.padding <- function(time){
    new_str <- paste(10000 + time, sep='')
    new_str <- substr(new_str, 2, nchar(new_str))
    new_str
}
```
## Loading and preprocessing the data
```{r}
actData <- read.csv("Data/activity.csv")
actData <- transform(actData, 
                     date=as.Date(date),
                     datetime=strptime(mapply(paste, date, time.padding(interval), collapse=" "),
                                       format="%Y-%m-%d %H%M"),
                     daytime=format(strptime(time.padding(interval),
                                             format="%H%M"), "%H:%M:%S"))
head(actData)
```


## What is mean total number of steps taken per day?
```{r}
stepsDay <- tapply(actData$steps[!is.na(actData$steps)],
                   actData$date[!is.na(actData$steps)], sum)
stepsDay <- as.vector(stepsDay)
stepsMean <- mean(stepsDay)
stepsMedian <- median(stepsDay)
qplot(stepsDay, xlab="Steps", ylab="Frequency") + geom_histogram(colour="black", 
                                                                 fill="white") +
    geom_vline(xintercept=stepsMean, color="red") + 
    geom_vline(xintercept=stepsMedian, color="green")

stepsMean
stepsMedian
```
The average and the median of the steps taken per day are `r format(stepsMean, nsmall=2, digits=3, scientific=F)` (steps/day) and `r format(stepsMedian, nsmall=2, digits=3, scientific=F)` (steps/day), respectively.

## What is the average daily activity pattern?
```{r}
stepsDayP <- tapply(actData$steps[!is.na(actData$steps)],
                    actData$daytime[!is.na(actData$steps)], mean)
dailyAct <- data.frame(time=strptime(names(stepsDayP), "%H:%M:%S"),
                       avg_steps=stepsDayP)

maxAvgInt <- dailyAct[dailyAct$avg_steps==max(dailyAct$avg_steps),]
ggplot(dailyAct, aes(x=time, y=avg_steps)) + geom_line() + xlab("Time") + ylab("Steps") + 
    scale_x_datetime(breaks=date_breaks("3 hours"), labels=date_format("%H:%M")) +
    geom_point(aes(x=maxAvgInt$time, y=maxAvgInt$avg_steps), col="red")
maxAvgInt
```
The 5-minute interval which contains the maximum number of steps occurs at `r rownames(maxAvgInt)`.
This interval is marked with a red dot in the plot above.

## Imputing missing values

There are `r sum(!complete.cases(actData))` missing values in the dataset.

Missing values from an interval are replaced by the mean for that interval as follows:
```{r}

noNAData <- transform(actData) 
for(i in (1:nrow(noNAData))[is.na(noNAData$steps)]){
    intv <- noNAData$daytime[i]
    noNAData$steps[i] <- dailyAct[rownames(dailyAct)==intv, c("avg_steps")]
}

```
The piece of code above takes any interval with missing values, and searches the calculated mean value in the
dataset stored in dailyAct, then it replaces de NA with the correspondin mean value.

Summary of the original dataset:
```{r}
summary(actData)
```


Summary of the dataset without missing values:
```{r}
summary(noNAData)
```

```{r}
stepsDay <- tapply(noNAData$steps[!is.na(noNAData$steps)],
                   noNAData$date[!is.na(noNAData$steps)], sum)
stepsDay <- as.vector(stepsDay)
stepsMean <- mean(stepsDay)
stepsMedian <- median(stepsDay)
qplot(stepsDay) + geom_histogram(colour="black", 
                                 fill="white") +
    geom_vline(xintercept=stepsMean, color="red") + 
    geom_vline(xintercept=stepsMedian, color="green") + xlab("Steps") + ylab("Frequency")

stepsMean
stepsMedian
```

The mean total number of steps per day remained the same while the median total number of steps per day 
changed with respect to the estimates calculated from the original dataset. In the histogram, only the bin 
containing the mean changed, the rest of the bins remained the same.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
noNAData <- transform(noNAData, daytype=factor(ifelse(weekdays(noNAData$date, abbreviate=T) %in% c("Sun", "Sat"),
                                                      "weekend", "weekday")))

weekdaysAct <- tapply(noNAData$steps[!is.na(noNAData$steps) & noNAData$daytype == "weekday"],
                      noNAData$daytime[!is.na(noNAData$steps) & "weekday" == noNAData$daytype], mean)
weekdaysAct <- data.frame(time=strptime(names(weekdaysAct), "%H:%M:%S"),
                          avg_steps=weekdaysAct, 
                          daytype=rep("weekday", nrow(weekdaysAct)))
weekendsAct <- tapply(noNAData$steps[!is.na(noNAData$steps) & "weekend" == noNAData$daytype],
                      noNAData$daytime[!is.na(noNAData$steps) & "weekend" == noNAData$daytype], mean)
weekendsAct <- data.frame(time=strptime(names(weekendsAct), "%H:%M:%S"),
                          avg_steps=weekendsAct, 
                          daytype=rep("weekend", nrow(weekendsAct)))
combinedDailyAct <- rbind(weekdaysAct, weekendsAct)
ggplot(combinedDailyAct, aes(x=time, y=avg_steps)) + geom_line() +facet_grid(daytype ~ .) + xlab("Time") + ylab("Steps") + 
    scale_x_datetime(breaks=date_breaks("3 hours"), labels=date_format("%H:%M")) 
```
